---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-",
  fig.cap=""
)
```
    
    
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/survminer)](http://cran.r-project.org/package=survminer)
    
  
# survminer: Survival Analysis and Visualization
    
The **survminer** R package provides functions for facilitating **survival analysis** and **visualization**. The current version contains the function **ggsurvplot()** for easily drawing beautiful survival curves using **ggplot2**. **ggsurvplot()** includes also some options for displaying the **p-value** and the **'number at risk' table**,  under the survival curves.
   
Find out more at http://www.sthda.com/english/wiki/survminer.
    
     
## Installation and loading
    
    
Install from CRAN as follow:
   
```{r, eval = FALSE}
install.packages("survminer")
```
    
    
Or, install the latest version from GitHub: 
    
   
```{r, eval = FALSE}
# Install
if(!require(devtools)) install.packages("devtools")
devtools::install_github("kassambara/survminer")
```

    
```{r}
# Loading
library("survminer")
```
    
    
    
## Getting started
     
     
```{r ggplot2-survival-plot, fig.show = "asis",  fig.height = 6, fig.width = 6}
#%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
# Example 1: Survival curve with one group
#%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

require("survival")
fit <- survfit(Surv(time, status) ~ 1, data = lung)
ggsurvplot(fit, color = "#2E9FDF")


#%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
# Example 2: Survival curves with two groups
#%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

# Fit survival curves
#++++++++++++++++++++++++++++++++++++
require("survival")
fit<- survfit(Surv(time, status) ~ sex, data = lung)

# Drawing survival curves
ggsurvplot(fit)

# Change font style, size and color
#++++++++++++++++++++++++++++++++++++
# Change only font size
ggsurvplot(fit, main = "Survival curve",
   font.main = 18,
   font.x =  16,
   font.y = 16,
   font.tickslab = 14)

# Change font size, style and color at the same time
ggsurvplot(fit, main = "Survival curve",
   font.main = c(16, "bold", "darkblue"),
   font.x = c(14, "bold.italic", "red"),
   font.y = c(14, "bold.italic", "darkred"),
   font.tickslab = c(12, "plain", "darkgreen"))

# Legend: title, labels and position
#++++++++++++++++++++++++++++++++++++

# Change the legend title and labels
ggsurvplot(fit, legend = "bottom", 
           legend.title = "Sex",
           legend.labs = c("Male", "Female"))

# Specify legend position by its coordinates
ggsurvplot(fit, legend = c(0.2, 0.2))


# format
#++++++++++++++++++++++++++++++++++++
# change line size --> 1
# Change line types by groups (i.e. "strata")
# and change color palette
ggsurvplot(fit,  size = 1,  # change line size
           linetype = "strata", # change line type by groups
           break.time.by = 250, # break time axis by 250
           palette = c("#E7B800", "#2E9FDF"), # custom color palette
           conf.int = TRUE, # Add confidence interval
           pval = TRUE # Add p-value
           )

# Use brewer color palette "Dark2"
ggsurvplot(fit, linetype = "strata", 
           conf.int = TRUE, pval = TRUE,
           palette = "Dark2")

# Use grey palette
ggsurvplot(fit, linetype = "strata", 
           conf.int = TRUE, pval = TRUE,
           palette = "grey")

# Add risk table
#++++++++++++++++++++++++++++++++++++

# Add risk table
ggsurvplot(fit, pval = TRUE, conf.int = TRUE,
           risk.table = TRUE)

# Customize the output and then print
res <- ggsurvplot(fit, pval = TRUE, conf.int = TRUE,
           risk.table = TRUE)
res$table <- res$table + theme(axis.line = element_blank())
res$plot <- res$plot + labs(title = "Survival Curves")
print(res)

# Change color, linetype by strata, risk.table color by strata
ggsurvplot(fit, 
           pval = TRUE, conf.int = TRUE,
           risk.table = TRUE, # Add risk table
           risk.table.col = "strata", # Change risk table color by groups
           lienetype = "strata", # Change line type by groups
           ggtheme = theme_bw(), # Change ggplot2 theme
           palette = c("#E7B800", "#2E9FDF"))

# Survival curve transformation
#++++++++++++++++++++++++++++++++++++
# Plot cumulative events
ggsurvplot(fit, conf.int = TRUE,
           palette = c("#FF9E29", "#86AA00"),
           risk.table = TRUE, risk.table.col = "strata",
           fun = "event")

# Plot the cumulative hazard function
ggsurvplot(fit, conf.int = TRUE, 
           palette = c("#FF9E29", "#86AA00"),
           risk.table = TRUE, risk.table.col = "strata",
           fun = "cumhaz")

# Arbitrary function
ggsurvplot(fit, conf.int = TRUE, 
          palette = c("#FF9E29", "#86AA00"),
           risk.table = TRUE, risk.table.col = "strata",
           pval = TRUE,
           fun = function(y) y*100)


#%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
# Example 3: Survival curve with multiple group
#%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

# Fit (complexe) survival curves
#++++++++++++++++++++++++++++++++++++

require("survival")
fit2 <- survfit( Surv(time, status) ~ rx + adhere,
    data = colon )

# Visualize
#++++++++++++++++++++++++++++++++++++

# Visualize: add p-value, chang y limits
# change color using brewer palette
ggsurvplot(fit2, pval = TRUE, 
           break.time.by = 400,
           risk.table = TRUE)

# Adjust risk table and survival plot heights
# ++++++++++++++++++++++++++++++++++++
# Risk table height
ggsurvplot(fit2, pval = TRUE,
          break.time.by = 400,
          risk.table = TRUE,
          risk.table.col = "strata",
          risk.table.height = 0.5, # Useful when you have multiple groups
          palette = "Dark2")
  
# Change legend labels
# ++++++++++++++++++++++++++++++++++++

ggsurvplot(fit2, pval = TRUE, 
           break.time.by = 400,
           risk.table = TRUE,
           risk.table.col = "strata",
           ggtheme = theme_bw(),
           legend.labs = c("A", "B", "C", "D", "E", "F"))
```

