---
title: "Competing risks: cumulative incidence, Gray's test, and the naive-KM trap"
author: "survminer"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Competing risks: cumulative incidence, Gray's test, and the naive-KM trap}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
has_tcr <- requireNamespace("tidycmprsk", quietly = TRUE)
```

When a subject can fail in more than one way -- relapse *or* death, one cancer
*or* another -- the events *compete*: the first one to happen removes the subject
from being at risk for the others. In that setting the ordinary Kaplan-Meier
estimate of "probability of event X" is biased upward, because it treats the
competing events as if they were merely censored. The right summary is the
**cumulative incidence function** (CIF), and survminer draws it with
`ggcompetingrisks()`.

```{r setup}
library(survminer)
library(survival)
```

We use the `mgus2` data (monoclonal gammopathy): the competing events are
progression to a **plasma-cell malignancy** (PCM) and **death** without
progression. The standard construction records, for each subject, the time to
the first event and which event it was:

```{r data}
mgus2$etime <- with(mgus2, ifelse(pstat == 1, ptime, futime))
mgus2$event <- with(mgus2, ifelse(pstat == 1, 1, 2 * death))
mgus2$event <- factor(mgus2$event, 0:2, labels = c("censor", "PCM", "death"))
mgus2$sex   <- factor(mgus2$sex,   labels = c("Female", "Male"))

table(mgus2$event)
```

```{r fallback, echo = FALSE, results = "asis", eval = !has_tcr}
cat("> **Note:** the cumulative-incidence figures below are built with the",
    "*tidycmprsk* package, which is not installed in this build, so they are",
    "not shown. Install it with `install.packages(\"tidycmprsk\")` to reproduce",
    "them.")
```

# Why not 1 − Kaplan-Meier?

A subject who dies without progressing can never progress afterwards. The
Kaplan-Meier estimate for PCM ignores that -- it censors the deaths and lets the
remaining subjects stand in for them, which inflates the estimated probability
of PCM to a value the data cannot support (in the limit, 1 − KM for every cause
can sum to more than 1). The cumulative incidence function counts a competing
event as leaving the risk set, so the incidences of all causes add up correctly.

# Cumulative incidence, the right way

`ggcompetingrisks()` accepts a `tidycmprsk::cuminc()` fit, which carries proper
pointwise confidence intervals. Fit the cumulative incidence by group and plot
one panel per cause:

```{r cif, eval = has_tcr}
cif <- tidycmprsk::cuminc(Surv(etime, event) ~ sex, data = mgus2)

ggcompetingrisks(cif, conf.int = TRUE, palette = "jco")
```

Each curve rises to the fraction of the cohort that has experienced that cause by
time *t*, with a shaded pointwise confidence band. Death (without progression) is
far more common than progression to PCM and accumulates faster in men;
progression to PCM is uncommon in either sex and, as Gray's test below shows, not
significantly different by sex.

# The naive-KM trap, made visible

The clearest way to see the bias is to overlay the naive 1 − KM estimate for one
cause on its cumulative incidence. `add.naive.km = TRUE` draws that contrast for
the chosen cause:

```{r naive, eval = has_tcr}
ggcompetingrisks(cif, add.naive.km = TRUE, cause = "PCM", palette = "jco")
```

The dashed line is 1 − KM for PCM; the solid line is its cumulative incidence.
The dashed curve sits well above the solid one and climbs steeply late in
follow-up: in women it reaches about **0.75** by the end of follow-up, against a
true cumulative incidence near **0.20** -- more than triple. That gap is the bias
the CIF avoids; reporting the dashed value would materially overstate a patient's
risk of progression.

# Testing between groups: Gray's test

Gray's test compares the cumulative incidence of a cause across groups (the
competing-risks analogue of the log-rank test). `pval = TRUE` annotates it per
cause:

```{r pval, eval = has_tcr}
ggcompetingrisks(cif, pval = TRUE, palette = "jco")
```

Here the incidence of death differs by sex (p = 0.00064) while progression to
PCM does not (p = 0.27). Gray's test is available on the `cmprsk` /
`tidycmprsk` path; an Aalen-Johansen (`survfit`) object carries no Gray's test.

# A number-at-risk table

`risk.table = TRUE` adds a per-group number-at-risk table beneath the plot,
locked to the same time axis. Note the layout changes: to share one time axis
with the table, the causes are drawn in a **single overlaid panel** -- each cause
is a colour and each group a **linetype** (solid Female, dashed Male) -- rather
than the separate panels above. (So here a dashed line marks the *male group*,
not the naive estimate of the earlier figure.) A subject who has had a competing
event counts as having left the risk set, matching the Kaplan-Meier convention
used elsewhere in survminer:

```{r risktable, eval = has_tcr, fig.height = 6}
ggcompetingrisks(cif, risk.table = TRUE, break.time.by = 60, palette = "jco")
```

# Where to next

- **Get started** -- presets, themes, and the composable curve.
- **Absolute effect measures** -- RMST, milestone, and landmark analysis.
- **Interrogating a Cox model** -- forest plots and proportional-hazards
  diagnostics.
