---
title: "Interrogating a Cox model: forest plots and proportional-hazards diagnostics"
author: "survminer"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Interrogating a Cox model: forest plots and proportional-hazards diagnostics}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5,
  dpi = 96
)
```

Once you have fitted a Cox model there are four questions a reader (or a
reviewer) will ask: *what does each covariate do?*, *is the treatment effect the
same in every subgroup?*, *does that effect survive adjustment?*, and *can I
trust the hazard ratio at all?* survminer answers each with a dedicated function
-- `ggforest()`, `ggforest_subgroup()`, `ggforest_models()`, and the
proportional-hazards diagnostic `ggcoxnph()`.

```{r setup}
library(survminer)
library(survival)

# Colon cancer adjuvant-therapy trial: overall survival (etype == 2), comparing
# observation with the Lev+5FU arm. A few covariates are given readable labels.
colon2 <- subset(colon, etype == 2 & rx %in% c("Obs", "Lev+5FU"))
colon2$rx     <- droplevels(colon2$rx)
colon2$sex    <- factor(colon2$sex,    labels = c("Female", "Male"))
colon2$differ <- factor(colon2$differ, labels = c("well", "moderate", "poor"))
colon2$agegrp <- cut(colon2$age, c(0, 50, 70, 100),
                     labels = c("<50", "50-70", ">70"))

cox <- coxph(Surv(time, status) ~ rx + sex + age + nodes + differ,
             data = colon2)
```

# One model's coefficients

`ggforest()` draws the hazard ratio and 95% confidence interval for every term
of a single Cox model, with the reference level, N, and p-value laid out beside
each row and a global model summary in the caption:

```{r forest, fig.height = 6}
ggforest(cox, data = colon2)
```

**Reading a forest plot.** The vertical reference line at **HR = 1** marks "no
effect": a box to its **left** (HR < 1) means that level lowers the hazard
relative to the reference, to its **right** (HR > 1) means it raises the hazard.
The box is the point estimate and the whiskers its 95% confidence interval; a
whisker that **crosses 1** is not statistically significant. The box *size*
grows with the precision of the estimate. (The same reference line appears in
every forest below.)

`ggforest()` has arguments to tailor the layout -- `columns` chooses which
statistic columns appear (any subset of `c("N", "hr", "ci", "p")`), and
`palette` colours the points and intervals (by variable, or a single accent)
instead of the default black.

# Is the effect the same in every subgroup?

A single model gives one averaged treatment effect. What clinicians publish is
the **treatment effect within each subgroup**, next to a test of whether the
effect actually differs across a subgroup's levels. `ggforest_subgroup()` draws
exactly that:

```{r subgroup, fig.height = 5, fig.width = 8.5}
ggforest_subgroup(
  cox, data = colon2,
  treatment = "rx",
  subgroups = c(Sex = "sex", "Age group" = "agegrp", Differentiation = "differ")
)
```

Read it in two parts, because it combines two different models:

- each **row** is the treatment hazard ratio estimated *within* that subgroup
  level (a Cox model refit on that subset), so you can see where the benefit is
  larger or smaller;
- each **header** carries one **interaction p-value** for the whole subgroup
  variable -- from a single model with a `treatment × subgroup` term (a
  likelihood-ratio test), *not* one p per level. A small interaction p is
  evidence of genuine effect modification; a large one says the apparent
  differences between levels are within noise.

This two-model structure is why a subgroup forest cannot be read off a single
`ggforest()`. Two details worth stating: the per-level hazard ratios are the
*treatment-only* effect within each subset (not adjusted for the other
covariates), and the `No. (%)` column counts the subgroup size -- so "Overall"
here is 619, the full cohort, which is larger than the 594 complete cases the
multivariable `ggforest()` above was fit on.

# One covariate across several models

When you add covariates to a model, the estimate for a key variable moves. The
classic reporting device -- unadjusted versus adjusted, or naive versus
time-varying -- puts that one covariate's hazard ratio on one row per model.
`ggforest_models()` does this:

```{r models, fig.height = 4}
cox_crude <- coxph(Surv(time, status) ~ rx, data = colon2)

ggforest_models(
  list("Unadjusted" = cox_crude, "Adjusted" = cox),
  term = "rx"
)
```

Here the treatment hazard ratio moves little after adjustment (0.69 unadjusted →
0.66 adjusted), consistent with limited confounding by the measured covariates
-- though part of the small shift is a change of sample, since the adjusted
model drops the few cases with missing covariates. The hazard ratio, confidence
interval and p-value are read from each model with a robust variance when one is
present, and a model missing the term is dropped with a warning.

# Can you trust the hazard ratio? The proportional-hazards check

Every hazard ratio above assumes **proportional hazards** -- that a covariate's
effect is a constant multiplier over all follow-up. When that fails, the single
HR is a time-average that can hide a treatment whose benefit grows, shrinks, or
reverses. `ggcoxnph()` gathers the standard non-proportional-hazards diagnostics
into one panel. We use the `veteran` lung-cancer trial and a performance-status
split (Karnofsky ≤ 60 vs > 60), a textbook example where the effect is strongly
time-varying:

```{r nph, fig.height = 6}
veteran$perf <- factor(ifelse(veteran$karno <= 60, "Poor", "Good"),
                       levels = c("Poor", "Good"))
cox_perf <- coxph(Surv(time, status) ~ perf, data = veteran)

ggcoxnph(cox_perf, variable = "perf", palette = "jco")
```

- **A. Complementary log-log survival** by group: parallel curves support
  proportional hazards; here they clearly converge, warning against it.
- **B. Scaled Schoenfeld residuals** with a smooth of β(t) and the
  Grambsch-Therneau p-value: a flat line (large p) is consistent with
  proportional hazards; here β(t) drifts and p = 0.00032, so PH is rejected.
- **C. Hazard-ratio trend** exp(β(t)): the *shape* of the effect over time --
  here the protective hazard ratio sits far below 1 early and drifts back toward
  it, so the benefit of good performance status is largest early and attenuates
  over follow-up (read it as shape, not magnitude).
- **D. RMST difference versus horizon**: the absolute benefit as the follow-up
  window lengthens -- a proportional-hazards-free view that previews the
  restricted mean survival time from the **Absolute effect measures** vignette.

# When the assumption fails

That is exactly this case: panel B rejects proportional hazards (p = 0.00032)
and panel A's curves converge, so a single hazard ratio for performance status
is not a faithful summary -- report an **absolute** effect measure instead. The
**Absolute effect measures** vignette covers restricted mean survival time,
milestone survival, and landmark analysis for this situation.

# Where to next

- **Get started** -- presets, themes, and the composable curve.
- **Absolute effect measures** -- RMST, milestone, and landmark analysis for
  when proportional hazards fails.
- **Competing risks** -- cumulative incidence when subjects can fail in more
  than one way.
