---
title: "Customization recipes"
author: "survminer"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Customization recipes}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

`ggsurvplot()` returns a compound object -- a list holding the survival
curve (`$plot`), the risk table (`$table`), the data actually plotted
(`$data.survplot`), and more -- and `surv_summary()` returns a plain data
frame with one row per event/censoring time (`time`, `surv`, `n.risk`,
`n.event`, `n.censor`, `strata`, ...). Because both are ordinary ggplot2 /
data-frame objects, a lot of "can survminer do X?" questions are answered by
*composing* on what these already give you rather than by a dedicated argument.

This page collects such recipes. Each is self-contained and uses the `lung`
data set from the survival package.

```{r setup}
library(survminer)
library(survival)
library(dplyr)
library(ggplot2)
```

# Dashed curves once few patients remain at risk

Some reporting guidelines ask that a survival curve be drawn with a **dashed
line beyond the point where few patients remain at risk** -- for example the
[guidelines for reporting mortality and morbidity after cardiac valve
interventions](https://pubmed.ncbi.nlm.nih.gov/18355567/) recommend a dashed
line once the number at risk falls below 10% of the initial cohort (survminer
issue [#559](https://github.com/kassambara/survminer/issues/559)).

`surv_summary()` carries `n.risk` per stratum, so you can find, for each curve,
the first time it drops below the threshold, split the curve into a solid
"dense" part and a dashed "sparse" tail, and draw the two with different
`linetype`s. Repeating the boundary point in the solid part makes the two step
segments meet with no gap.

```{r dashed-helper}
# KM curves that switch solid -> dashed once the number at risk drops below
# `frac` of each group's initial cohort (default 10%).
ggsurv_dashed_below <- function(fit, data, frac = 0.10, ...) {
  ss <- surv_summary(fit, data = data) %>%
    group_by(strata) %>%
    mutate(onset = { s <- n.risk < frac * max(n.risk)
                     if (any(s)) min(time[s]) else Inf },
           phase = ifelse(time >= onset, "sparse", "dense")) %>%
    ungroup()
  # repeat the first sparse point in the solid phase so the two steps join
  bridge <- ss %>% filter(phase == "sparse") %>%
    group_by(strata) %>% slice_min(time, n = 1) %>% ungroup() %>%
    mutate(phase = "dense")

  ggplot(bind_rows(ss, bridge), aes(time, surv, colour = strata)) +
    geom_step(aes(linetype = phase), linewidth = 0.9) +
    geom_point(data = filter(ss, n.censor > 0), shape = 3, size = 2,
               show.legend = FALSE) +
    scale_linetype_manual(values = c(dense = "solid", sparse = "dashed"),
                          breaks = "sparse",
                          labels = sprintf("< %g%% at risk", 100 * frac),
                          name = NULL) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = "Time", y = "Survival probability", ...) +
    theme_survminer()
}
```

```{r dashed-plot}
fit <- survfit(Surv(time, status) ~ sex, data = lung)
ggsurv_dashed_below(fit, lung, frac = 0.10) +
  scale_colour_discrete(name = "Sex", labels = c("Male", "Female"))
```

Each curve turns dashed at its own crossing point. A few adjustments:

- The threshold is each group's own initial cohort (`max(n.risk)` per stratum).
  To key it to the pooled study size instead, replace `max(n.risk)` with the
  overall starting `n` and every curve dashes at the same count.
- `frac` is an argument, so the same helper produces the guideline's 10% cutoff
  or any other fraction.
- This rebuilds the line layer to control `linetype`, so it does not go through
  `ggsurvplot()` directly. If you also want the number-at-risk table underneath,
  build it separately with `ggrisktable()` on the same fit and stack the two
  (for example with `ggpubr::ggarrange(..., ncol = 1, heights = c(3, 1))`).

# Step-shaped legend keys

By default the legend key is a straight horizontal line. Some publications draw
the key as a small step that echoes the Kaplan-Meier shape (survminer issue
[#537](https://github.com/kassambara/survminer/issues/537)). `ggsurvplot()`
returns the curve as `$plot`, whose first layer draws the survival lines, so you
can swap that layer's legend key for a custom step glyph -- overriding only the
key, leaving the curve drawing untouched.

```{r step-key-helper}
library(grid)

# Draw the legend key of a ggsurvplot's curves as a small step (stair) that
# echoes the Kaplan-Meier shape, instead of the default straight line.
step_legend_key <- function(p) {
  `%||%` <- function(x, y) if (is.null(x)) y else x   # base R has this since 4.4
  draw_key_step <- function(data, params, size) {
    grid::linesGrob(
      x = c(0, 0.45, 0.45, 1), y = c(0.25, 0.25, 0.7, 0.7),
      gp = grid::gpar(col = data$colour %||% "black",
                      lwd = (data$linewidth %||% data$size %||% 0.75) * .pt,
                      lty = data$linetype %||% 1, lineend = "butt"))
  }
  # the survival curves are the first layer; subclass its geom to override ONLY
  # the legend key, so the curve drawing itself is unchanged
  g <- p$plot$layers[[1]]$geom
  p$plot$layers[[1]]$geom <-
    ggplot2::ggproto(paste0(class(g)[1], "Key"), g, draw_key = draw_key_step)
  p
}
```

```{r step-key-plot}
fit2 <- survfit(Surv(time, status) ~ sex, data = lung)
p <- ggsurvplot(fit2, data = lung, palette = c("#E7302A", "#2E43F3"),
                legend.title = "Group", legend.labs = c("High", "Low"))
step_legend_key(p)
```

Adjust the `x`/`y` coordinates inside `draw_key_step()` to change the step
shape (for example a taller riser, or a two-step stair).

# Mark the median survival and its confidence interval

`surv.median.line = "hv"` already drops a line at each group's median survival.
To also show the *confidence interval* of the median (survminer issue
[#345](https://github.com/kassambara/survminer/issues/345)), use
`surv_median()`, which returns `median`, `lower` and `upper` per group, and add
a horizontal band at survival = 0.5.

```{r median-ci}
add_median_ci <- function(p, fit, y = 0.5) {
  med  <- surv_median(fit)
  # one colour per group, read from the drawn curve layer (no internal helpers)
  b     <- ggplot2::ggplot_build(p$plot)$data[[1]]
  cols  <- b$colour[match(sort(unique(b$group)), b$group)]
  # drop groups whose median is not reached (surv_median() returns NA there)
  keep  <- !is.na(med$median)
  med   <- med[keep, ]
  cols  <- cols[keep]
  p$plot +
    geom_segment(data = med, aes(x = lower, xend = upper, y = y, yend = y),
                 colour = cols, linewidth = 0.9, inherit.aes = FALSE) +
    geom_point(data = med, aes(x = median, y = y),
               colour = cols, size = 2.4, inherit.aes = FALSE)
}
```

```{r median-ci-plot}
fit3 <- survfit(Surv(time, status) ~ sex, data = lung)
p <- ggsurvplot(fit3, data = lung, palette = c("#E7302A", "#2E43F3"),
                surv.median.line = "hv",
                legend.labs = c("Male", "Female"), legend.title = "Sex")
p$plot <- add_median_ci(p, fit3)
p
```

The horizontal bar spans the 95% CI of the median survival time
(`surv_median()`'s `lower`/`upper`); the point marks the median itself. A group
whose survival never drops to 0.5 has no median (`surv_median()` returns `NA`),
so it is simply left without a bar.

# Plot math (expressions) in the legend

`legend.labs = c(expression(...))` alone does not render as math, and `+` does
not apply to the compound `ggsurvplot` object. Add the scale to `$plot` (a plain
ggplot) instead, supplying the same palette so the colours are unchanged
(survminer issue [#350](https://github.com/kassambara/survminer/issues/350)).

```{r plotmath-legend}
pal <- c("#E7302A", "#2E43F3")
p <- ggsurvplot(fit3, data = lung, palette = pal, legend.title = "Group")
p$plot <- p$plot +
  scale_colour_manual(values = pal,
                      labels = c(expression(x^2 ~ y^2), expression(alpha[i] >= beta)),
                      name = "Group")
p
```

(The "Scale for colour is already present" message is expected -- we are
replacing survminer's default colour scale with one that carries the math
labels.)

# One shared legend across a grid of plots

Several separate survival analyses that use the same grouping are often shown in
a grid with a single common legend (survminer issue
[#340](https://github.com/kassambara/survminer/issues/340)). Pull each
`ggsurvplot`'s `$plot` and arrange them with `ggpubr::ggarrange()`, which draws
one shared legend and keeps every panel the same size.

```{r shared-legend, fig.height = 6}
colon2  <- colon[!is.na(colon$sex), ]
subsets <- split(colon2, colon2$extent)
subsets <- subsets[sapply(subsets, nrow) > 30]
subsets <- subsets[seq_len(min(4, length(subsets)))]

plots <- lapply(names(subsets), function(nm) {
  d   <- subsets[[nm]]
  fit <- survfit(Surv(time, status) ~ sex, data = d)
  ggsurvplot(fit, data = d, legend.title = "Sex",
             legend.labs = c("Male", "Female"), title = paste("extent", nm))$plot
})

ggpubr::ggarrange(plotlist = plots, ncol = 2, nrow = 2,
                  common.legend = TRUE, legend = "bottom")
```
