---
title: "Customization recipes"
author: "survminer"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Customization recipes}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5
)
```

`ggsurvplot()` returns a compound object -- a list holding the survival
curve (`$plot`), the risk table (`$table`), the data actually plotted
(`$data.survplot`), and more -- and `surv_summary()` returns a plain data
frame with one row per event/censoring time (`time`, `surv`, `n.risk`,
`n.event`, `n.censor`, `strata`, ...). Because both are ordinary ggplot2 /
data-frame objects, a lot of "can survminer do X?" questions are answered by
*composing* on what these already give you rather than by a dedicated argument.

This page collects such recipes. Each is self-contained and uses the `lung`
data set from the survival package.

```{r setup}
library(survminer)
library(survival)
library(dplyr)
library(ggplot2)
```

# Dashed curves once few patients remain at risk

Some reporting guidelines ask that a survival curve be drawn with a **dashed
line beyond the point where few patients remain at risk** -- for example the
[guidelines for reporting mortality and morbidity after cardiac valve
interventions](https://pubmed.ncbi.nlm.nih.gov/18355567/) recommend a dashed
line once the number at risk falls below 10% of the initial cohort (survminer
issue [#559](https://github.com/kassambara/survminer/issues/559)).

`surv_summary()` carries `n.risk` per stratum, so you can find, for each curve,
the first time it drops below the threshold, split the curve into a solid
"dense" part and a dashed "sparse" tail, and draw the two with different
`linetype`s. Repeating the boundary point in the solid part makes the two step
segments meet with no gap.

```{r dashed-helper}
# KM curves that switch solid -> dashed once the number at risk drops below
# `frac` of each group's initial cohort (default 10%).
ggsurv_dashed_below <- function(fit, data, frac = 0.10, ...) {
  ss <- surv_summary(fit, data = data) %>%
    group_by(strata) %>%
    mutate(onset = { s <- n.risk < frac * max(n.risk)
                     if (any(s)) min(time[s]) else Inf },
           phase = ifelse(time >= onset, "sparse", "dense")) %>%
    ungroup()
  # repeat the first sparse point in the solid phase so the two steps join
  bridge <- ss %>% filter(phase == "sparse") %>%
    group_by(strata) %>% slice_min(time, n = 1) %>% ungroup() %>%
    mutate(phase = "dense")

  ggplot(bind_rows(ss, bridge), aes(time, surv, colour = strata)) +
    geom_step(aes(linetype = phase), linewidth = 0.9) +
    geom_point(data = filter(ss, n.censor > 0), shape = 3, size = 2,
               show.legend = FALSE) +
    scale_linetype_manual(values = c(dense = "solid", sparse = "dashed"),
                          breaks = "sparse",
                          labels = sprintf("< %g%% at risk", 100 * frac),
                          name = NULL) +
    coord_cartesian(ylim = c(0, 1)) +
    labs(x = "Time", y = "Survival probability", ...) +
    theme_survminer()
}
```

```{r dashed-plot}
fit <- survfit(Surv(time, status) ~ sex, data = lung)
ggsurv_dashed_below(fit, lung, frac = 0.10) +
  scale_colour_discrete(name = "Sex", labels = c("Male", "Female"))
```

Each curve turns dashed at its own crossing point. A few adjustments:

- The threshold is each group's own initial cohort (`max(n.risk)` per stratum).
  To key it to the pooled study size instead, replace `max(n.risk)` with the
  overall starting `n` and every curve dashes at the same count.
- `frac` is an argument, so the same helper produces the guideline's 10% cutoff
  or any other fraction.
- This rebuilds the line layer to control `linetype`, so it does not go through
  `ggsurvplot()` directly. If you also want the number-at-risk table underneath,
  build it separately with `ggrisktable()` on the same fit and stack the two
  (for example with `ggpubr::ggarrange(..., ncol = 1, heights = c(3, 1))`).
