---
title: "Absolute effect measures: RMST, milestone and landmark analysis"
author: "survminer"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Absolute effect measures: RMST, milestone and landmark analysis}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

The log-rank p-value tells you *whether* two survival curves differ; the hazard
ratio summarises the difference as a single ratio -- but only if the hazards
stay proportional. When curves cross, converge, or separate late, a reviewer
increasingly asks for an **absolute** effect measure in the units of the study:
months of life gained, or the survival probability at a clinically meaningful
time. This article covers the three survminer draws for you: restricted mean
survival time, milestone survival, and landmark analysis.

```{r setup}
library(survminer)
library(survival)

# Readable arm names for every figure.
lung$sex <- factor(lung$sex, labels = c("Male", "Female"))
fit <- survfit(Surv(time, status) ~ sex, data = lung)
```

# Why a single hazard ratio may not be enough

A Cox hazard ratio assumes the effect of a group is a constant multiplier of the
hazard over all follow-up. Under non-proportional hazards that single number
averages together periods where one arm does better and periods where it does
worse, and it is no longer a faithful summary. The absolute measures below make
no proportional-hazards assumption -- they read the difference directly off the
Kaplan-Meier curves.

# Restricted mean survival time (RMST)

The restricted mean survival time up to a horizon **τ** is the *area under the
survival curve* from 0 to τ -- the average survival time a subject can expect
over the first τ units of follow-up. The **difference** in RMST between two arms
is an absolute benefit measured in time units (here, days), and unlike the
hazard ratio it needs no proportional-hazards assumption.

`ggrmst()` shades the area between the two curves up to τ and annotates the
difference with its confidence interval and p-value:

```{r rmst}
ggrmst(fit, data = lung, tau = 500, palette = "jco",
       legend.labs = c("Male", "Female"))
```

Here the curves do not cross, so the shaded region *equals* the ΔRMST: over the
first 500 days, women survive on average about 82 **days** longer than men.
Because RMST is an area, it stays well defined even when the curves cross --
exactly the case where a hazard ratio struggles. (When the curves *do* cross,
the shaded band shows the *total* separation and the reported difference, which
nets the two directions, can be smaller than the band.)

## Reading and reporting RMST

`ggrmst_difference()` returns the numbers behind the figure: the per-arm RMST
with its standard error and confidence interval, and the between-arm difference
with its confidence interval and p-value.

```{r rmst-table}
ggrmst_difference(fit, data = lung, tau = 500)
```

**Choosing τ.** RMST is only defined up to a horizon where all arms still have
subjects at risk, so τ must lie within the follow-up of every arm. If you omit
`tau`, survminer uses the largest commonly admissible time; if you pass a τ
beyond an arm's support, it stops with a clear message rather than extrapolating.
Report τ alongside the estimate -- "the RMST difference at τ = 500 days" -- since
the value depends on it.

# Milestone (fixed-time) survival

A milestone is the survival probability at a fixed clinical timepoint --
one-year survival, two-year survival -- and the quantity a reader often wants is
the **between-arm difference** at that time, with a confidence interval.
`ggmilestone()` marks each timepoint and reports that difference:

```{r milestone, fig.height = 5.5}
ggmilestone(fit, data = lung, milestone.times = c(180, 360, 540),
            palette = "jco", legend.labs = c("Male", "Female"))
```

Each dashed line is a milestone; the caption gives, per timepoint, the survival
in each arm and their difference with a 95% confidence interval and p-value. The
difference is reported as Male − Female, so the negative values mean women do
better at every timepoint -- consistent with the positive Female − Male RMST
above. The full table is also attached to the plot:

```{r milestone-table}
p <- ggmilestone(fit, data = lung, milestone.times = c(180, 360, 540))
attr(p$plot, "milestone.table")
```

A milestone that falls beyond an arm's follow-up is returned as `NA` with a
warning rather than silently dropped.

# Landmark analysis and immortal-time bias

When the groups you compare are defined by something that happens *after*
baseline -- responded to treatment by week 8, received a transplant, achieved
remission -- a naive Kaplan-Meier from time zero is biased. To be classified in
the "responder" group a subject must first *survive* long enough to respond, so
that group is credited with "immortal" time it was guaranteed to survive. This
is **immortal-time (guarantee-time) bias**, and it makes the post-baseline group
look artificially better.

The standard remedy is a **landmark analysis**: pick a landmark time `L` after
which group membership is fixed, keep only subjects still event-free and at risk
at `L`, and reset the clock to `time - L`. `gglandmark()` does exactly this:

`lung` has no post-baseline grouping variable, so we apply the landmark to `sex`
only to show the mechanics -- being a fixed *baseline* covariate, `sex` does not
itself suffer immortal-time bias; in practice the grouping would be the
post-baseline event (responder, transplant, remission).

```{r landmark}
lm <- gglandmark(fit, data = lung, landmark.time = 180, palette = "jco",
                 legend.labs = c("Male", "Female"))
# the function's long auto-subtitle is replaced by a short one; the details
# are in the text below
lm$plot <- lm$plot + ggplot2::labs(subtitle = "Landmark at day 180")
lm
```

The curves now start at `L` (shown as time 0 on a "Time since landmark" axis),
restricted to the subjects still at risk then -- at day 180, 159 subjects remain
at risk, and one subject with an event exactly at `L` is excluded so each curve
correctly begins at S = 1. Any log-rank p-value or risk table you request is
computed on this landmark cohort.

Use a landmark analysis whenever the grouping variable is measured after
follow-up begins; choose `L` from clinical reasoning (when the classifying event
has typically occurred), not from the data.

# Where to next

- **Get started** -- presets, themes, and the composable curve.
- **Interrogating a Cox model** -- forest plots and proportional-hazards
  diagnostics (`ggcoxnph()`), the diagnostic that tells you *when* to reach for
  the absolute measures above.
- **Competing risks** -- cumulative incidence when subjects can fail in more
  than one way.
