---
title: "Get started: from data to a publication-ready figure"
author: "survminer"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{Get started: from data to a publication-ready figure}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  collapse = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5,
  dpi = 96
)
```

survminer turns a survival fit into a figure you can put straight into a
manuscript. This page is the three-minute tour: fit a curve, get a
publication-ready figure with one argument, adjust the look, report the numbers
a reviewer expects, and build on the plot when you need something bespoke.

```{r setup}
library(survminer)
library(survival)

# Label the sex factor once, so every figure shows readable arm names
# (rather than "sex=1" / "sex=2").
lung$sex <- factor(lung$sex, labels = c("Male", "Female"))
```

Every example uses the `lung` data set (NCCTG lung cancer) from the survival
package, so you can paste and run any chunk as-is.

# Three lines to a survival curve

Fit a Kaplan-Meier curve with `survival::survfit()`, then draw it with
`ggsurvplot()`. That is the whole minimal workflow:

```{r first-curve}
fit <- survfit(Surv(time, status) ~ sex, data = lung)
ggsurvplot(fit, data = lung)
```

`ggsurvplot()` already gives you a coloured curve per group, a censoring
legend, and clean axis labels. From here every element -- confidence bands, a
p-value, a number-at-risk table -- is one more argument.

# One argument to a publication-ready figure

Instead of setting a dozen arguments by hand, pass `preset = "publication"`. It
bundles the evidence a survival figure is expected to carry -- confidence
bands, the log-rank p-value, median-survival crosshairs, a number-at-risk
table, a colourblind-friendly palette, and a percent y-axis:

```{r publication, fig.height = 6.5}
ggsurvplot(fit, data = lung, preset = "publication")
```

Everything in that figure maps to an argument you could have set yourself:

| Visible element                    | Argument the preset set          |
|------------------------------------|----------------------------------|
| Shaded confidence bands            | `conf.int = TRUE`                |
| p-value, labelled "Log-rank"       | `pval = TRUE, pval.method = TRUE`|
| Dashed median-survival crosshairs  | `surv.median.line = "hv"`        |
| Number-at-risk table below         | `risk.table = TRUE`              |
| Colourblind-friendly colours       | `palette = "jco"`                |
| Survival as a percentage           | `surv.scale = "percent"`         |

The preset is a starting point, not a straitjacket -- the next section shows
how your own arguments override it.

# Your arguments always win

A preset only fills in the arguments you did **not** supply. Anything you pass
explicitly wins, so you can take the publication bundle and change just one
thing -- here, the palette:

```{r override, fig.height = 6.5}
ggsurvplot(
  fit, data = lung,
  preset  = "publication",
  palette = c("#E7B800", "#2E9FDF")   # your palette overrides the preset's "jco"
)
```

The default is `preset = "none"`, which changes nothing -- existing code keeps
its exact output.

# Choosing a look

There are four presets. `"publication"` is the full evidence panel above;
`"classic"` is a greyscale-safe textbook look (one line type per group);
`"minimal"` is an airy, stripped-back curve; `"presentation"` uses bold, large
type for slides. Compare them on the same fit:

To compare them like-for-like we override two preset defaults so every panel is
the same height -- turning the risk table and the median lines off (another
illustration of "your arguments win") -- and look only at the styling:

```{r presets-grid, fig.width = 9.5, fig.height = 6.5}
looks <- c("classic", "minimal", "presentation", "publication")
plots <- lapply(looks, function(p)
  ggsurvplot(fit, data = lung, preset = p, title = p,
             risk.table = FALSE, surv.median.line = "none"))
arrange_ggsurvplots(plots, ncol = 2, nrow = 2)
```

Pick `"publication"` or `"classic"` for a paper, `"presentation"` for a talk,
`"minimal"` when the curve is one panel of a larger figure.

# Reusable themes

A preset bundles the *look* **and** the evidence layers (bands, p-value, table);
a theme restyles only the *look*. The looks above are also available as
standalone ggplot2 themes you can apply to any survminer plot (or any ggplot)
through `ggtheme`, without a preset: `theme_surv_classic()`,
`theme_surv_minimal()`, and `theme_surv_bold()`.

```{r themes}
ggsurvplot(
  fit, data = lung,
  conf.int = TRUE,
  palette  = "jco",
  ggtheme  = theme_surv_bold()
)
```

# The numbers every KM paper reports

Two numbers belong in the caption of almost every Kaplan-Meier figure: the
**median survival** per group, and the **median follow-up** (the typical
observation time -- estimated by the reverse Kaplan-Meier method, which swaps
the event and censoring roles).

```{r numbers}
# Median survival per group (with 95% CI)
surv_median(fit)

# Median follow-up per group (reverse Kaplan-Meier)
surv_median_followup(fit, data = lung)
```

Both return tidy data frames, so you can drop them into a table or paste the
values into a figure caption.

# Building on the curve

`ggsurvplot()` returns a compound object (the curve in `$plot`, the risk table
in `$table`, and more), which is ideal for a finished figure but is not a plain
ggplot you can freely add layers to. When you want to compose your own
annotations, ask for a bare ggplot with `output = "ggplot"`:

The bare curve does not carry the compound object's risk table or the preset
styling -- you add exactly the layers you want:

```{r composable}
p <- ggsurvplot(fit, data = lung, palette = "jco", output = "ggplot")

# it is a genuine ggplot, so ordinary ggplot2 verbs work
p +
  ggplot2::geom_hline(yintercept = 0.5, linetype = "dashed", colour = "grey50") +
  ggplot2::annotate("text", x = 0, y = 0.53, hjust = 0, size = 3,
                    label = "median survival") +
  ggplot2::labs(title = "Add any layer you like")
```

Because it is a real ggplot, `+ geom_*`, `facet_wrap()`, and `ggsave()` all
behave normally -- the bare curve is the surface to build RMST shading, a
competing-risks overlay, or a parametric fit on (see the other articles).

# Saving your figure

`ggsave()` works directly on the object `ggsurvplot()` returns -- the curve and
its risk table are saved together, correctly aligned:

```{r save, eval = FALSE}
fig <- ggsurvplot(fit, data = lung, preset = "publication")
ggplot2::ggsave("survival.png", fig, width = 7, height = 6.5, dpi = 300)
```

For a journal, a width of 7 inches at 300 dpi is a safe starting point.

# Where to next

- **Absolute effect measures beyond the p-value** -- restricted mean survival
  time (RMST), milestone survival, and landmark analysis.
- **Interrogating a Cox model** -- forest plots (`ggforest()`,
  `ggforest_models()`, `ggforest_subgroup()`) and proportional-hazards
  diagnostics.
- **Competing risks** -- cumulative incidence done right, with Gray's test and
  the naive-Kaplan-Meier contrast.
- **Customization recipes** -- compose on the objects survminer returns to
  solve one-off requests.
